---
title: "Week 8 Demonstration"
format: html
editor: visual
---

## Set up

```{r}
#| message: FALSE
library(fpp3)
library(tidyverse)
library(slider)
library(gridExtra)
```

## 1. Residual analysis

a.  What is the difference between innovation residuals and regular residuals? Why do we perform residual analysis on the former rather than the latter?
b.  Starting with the following code snippet, fit several models to the Australian takeaway turnover time series and analyze their residuals. Which models have a good fit? Compare this with their CV error.

```{r}
takeaway <- aus_retail |>
  filter(Industry == "Takeaway food services") |>
  summarise(Turnover = sum(Turnover))
```

## 2. ACF for AR(2)

a.  Using `ARMAacf`, plot the ACF for AR(2) models with the following coefficients:

-   $(X_t)$: $\phi_1 = 1.5$, $\phi_2 = -0.55$
-   $(Y_t)$: $\phi_1 = 1.5$, $\phi_2 = -0.75$

b.  What is the qualitative behavior of the ACFs?
c.  From the ACF plots, try to guess what patterns you may observe from the sample trajectories.

## 3. Sample trajectories

a.  Using `arima.sim()`, draw sample trajectories from the AR(2) models in Q2.
b.  Fit AR(2) models to these generated time series using `ARIMA()`.
c.  What are the fitted coefficients?
d.  Plot the forecast curves.

## 4. AR(2) solution

Consider the AR(2) equation $$
X_t = 1.5 X_{t-1} - 0.75 X_{t-2} + W_t.
$$

a.  What is the autoregressive polynomial?

b.  What is the formula for the stationary solution?

## 5. Nonlinear autoregressive models

Fit a nonlinear AR model on the `globtemp` dataset. Compute its sum of squared residuals and compare it to that of a linear AR model.
