---
title: "Week 12 Demonstration"
format: pdf
editor: visual
---

## Set up

```{r}
#| message: FALSE
library(fpp3)
library(tidyverse)
library(slider)
library(gridExtra)
```

## 1. Finding code

Look at the CRAN task views page to find what packages are available for state space models.

## 2. Local linear trend model

A local linear trend model is defined by the equations $$
X_t = X_{t-1} + \alpha + W_t,
$$ $$
Y_t = X_t + V_t.
$$ We can interpret this as the time series $(Y_t)$ comprising a "trend" component, modeled as a random walk with drift, and a random component $(V_t)$.

a.  How can one write this in the form of a state space model?

b.  Simulate a time series of length 50 from this model and plot the results.

c.  Use the functions `SSModel()` and `SSMcustom()` in the `KFAS` package to specify a state space model. Read the documentation on how to do this.

d.  Use `KFS()` to performing filtering, smoothing, and one-step-ahead forecasting for $(X_t)$. Read the documentation to understand the output of the function.

e.  Plot the three time series against the true $(X_t)$ values and the observed values $(Y_t)$. Also plot level 0.95 confidence bands.

f.  What are the differences between the three time series?

## 3. Global temperature

Consider the two time series `gtemp_land` and `gtemp_ocean` discussed in the lecture video. Load, and normalize these time series via the following code snippet.

```{r}
library(astsa)
temp_dat <- tibble(ocean = gtemp_ocean / sd(gtemp_ocean), 
                   land = gtemp_land / sd(gtemp_land)) 
```

a.  What is an appropriate state space model for these time series?

b.  What are the parameters to be estimated?

c.  Fit the parameters of the model using `fitSSM()`.

d.  Use the fitted model to perform smoothing, and plot the resulting values, together with a level 0.95 confidence band and the two original time series.

## 4. Biomarker monitoring

Load the biomarker dataset using the following code snippet.

```{r}
biomarker <- 
    tibble(Day = 1:length(WBC), 
           WBC = WBC, 
           PLT = PLT, 
           HCT = HCT) |> 
    as_tsibble(index = Day) |> 
    mutate(across(everything(), ~replace(., . == 0, NA)))
```

a.  What is an appropriate state space model for these time series?

b.  What are the parameters to be estimated?

c.  Fit the parameters of the model using `fitSSM()`.

d.  Use the fitted model to perform filtering, and plot the resulting values, together with a level 0.95 confidence band and the two original time series.

e.  Interpret the fitted parameter values for $\Phi$.

f.  Why do our results look different compared to those in Figure 6.6 in Shumway and Stoffer (page 315)?
