---
title: "Week 5 Demonstration"
format: html
editor: visual
---

## Set up

```{r}
#| message: FALSE
library(fpp3)
library(tidyverse)
library(slider)
```

## 1. White noise and ACF plots

Create a time series of length 200 comprising i.i.d. standard Gaussian measurements. Make a time plot of the time series as well as an ACF plot. Compare this with the remainder term from an STL decomposition of `aus_arrivals_jap`. What do you observe?

## 2. Summary statistics

Glance through the summary statistics described in [Chapter 4.4](https://otexts.com/fpp3/other-features.html) of Hyndman and Athanasopoulos (2021). What do the following summary statistics compute?

-   Strength of trend
-   Strength of seasonality
-   `shift_level_max`
-   `shift_level_index`
-   `var_tiled_var`

When would they be useful?

## 3. EDA with summary statistics

The follow code snippet generates summary statistics for the various time series in the `tourism` dataset.

```{r}
#| eval: FALSE
tourism_features <- tourism |>
  features(Trips, feature_set(pkgs = "feasts"))
```

Make a scatter plot of seasonal strength against trend strength and color the points according to `Purpose`. What can you tell from the plot?

Perform PCA on the `tourism_features` dataframe and make a scatter plot of the first two PCs. What do you observe?

## 4. Seasonal periods and frequencies

Consider the time series $(y_t)$ generated using the following code snippet.

```{r}
y_dat <- tibble(t = 1:100,
       y = t %% 10) |>
  as_tsibble(index = t)
```

Plot a periodogram for $(y_t)$. What periods do the peaks correspond to? Why are there multiple peaks when $(y_t)$ was defined using a seasonal period of 10?

## 5. Periodogram and filters

Consider tourist arrivals to Australia from the UK, contained in the `aus_arrivals` dataset.

a.  Plot a periodogram for the time series of arrivals from UK. Where are the peaks?

b.  Apply a moving average of window size 4 to the time series. What happens to the periodogram?

c.  Apply a difference transformation to the time series. What happens to the periodogram?
