---
title: "ST5209/X Assignment 4"
format: pdf
editor: visual
author: "Due 31 Mar, 11.59pm"
---

## Set up

1.  Make sure you have the following installed on your system: $\text{\LaTeX}$, R4.2.2+, RStudio 2023.12+, and Quarto 1.3.450+.
2.  Pull changes from the course [repo](https://github.com/yanshuotan/st5209-2024).

## Submission

1.  Render the document to get a .pdf printout.
2.  Submit both the .qmd and .pdf files to Canvas.

```{r}
#| message: FALSE
#| echo: FALSE
library(tidyverse)
library(fpp3)
```

## 1. AR polynomial

Consider the AR(2) model $$
X_t = 4 + 0.5 X_{t-1} - 0.25 X_{t-2} + W_t.
$$ {#eq-ar2}

a.  What is the autoregressive polynomial?

b.  What are its roots?

c.  Is this model causal? Why?

d.  Given the representation $$
    X_t = \mu + \sum_{j=0}^\infty \psi_j W_{t-j},
    $$ solve for $\psi_0, \psi_1, \psi_2, \psi_3$. *Hint: Follow the strategy discussed in `linear_process_coefs.qmd`.*

## 2. Likelihood

Consider the AR(1) model $$
X_t = 1-0.6X_{t-1} + W_t,
$$ where $W_t \sim WN(0, 0.25)$ is Gaussian white noise. We are given observations $x_1 = 0.2, x_2 = -0.3, x_3 = 0.4$.

a.  What is the mean of $X_t$?

b.  Write the conditional likelihood of this model, when conditioning on the value of $x_1$.

## 3. Reversibility

a.  Create a sample trajectory of length $n = 200$ from the AR(2) model from Problem 1 using `arima.sim()`.

b.  Reverse the time index of the vector you obtain using `rev()`.

c.  Fit an AR(2) model to the reversed time series using `fable`. Hint: You may use the code snippet `model(AR(X ~ order(2)))`.

d.  Inspect the model parameters using `tidy()`. Why are they similar to the those in @eq-ar2?

e.  Make a forecast with $h = 10$. What does this correspond to in terms of the original time series?

## 4. Yule-Walker

a.  Write the Yule-Walker equations for the AR(2) model from Problem 1.

b.  Arrange the equations in the following matrix form (i.e. fill in the missing entries): $$
    \left[\begin{matrix}     ? & ? & ?  \\ ? & ? & ? \\
    ? & ? & ?
    \end{matrix}\right] \left[\begin{matrix}     \gamma(0) \\ \gamma(1) \\ \gamma(2)
    \end{matrix}\right] = \left[\begin{matrix}     \sigma^2 \\ 0 \\ 0
    \end{matrix}\right].
    $$

c.  Solve the system from part b) for $\gamma(0), \gamma(1), \gamma(2)$ numerically using `solve()`.

d.  Given $$
    \Gamma_2 = \left[\begin{matrix}     \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0)
    \end{matrix}\right],
    $$ what is the top left entry of $\Gamma_2^{-1}$ in terms of $\gamma(0)$ and $\gamma(1)$?

e.  Write a 95% confidence interval for $\phi_1$ using your answers for d) and 3d).

## 5. Recursive forecasting for ARMA(1, 1)

Consider the $ARMA(1,1)$ model $$
X_t - 0.5X_{t-1} = W_t + 0.5W_{t-1}.
$$ In this question, we will investigate recursive forecasting. The following code snippet generates a sequence of length $n=50$ drawn from the above model.

```{r}
set.seed(5209)
n <- 50
wn <- rnorm(n)
xt <- arima.sim(model = list(ar = 0.5, ma = 0.5), innov = wn, n = n)
```

a.  Fill in the following code snippet using equation (11.14) to generate a sequence `wn_hat`.

```{r}
wn_hat <- rep(0, n)
wn_hat[[1]] <- xt[[1]]
for (i in 2:n) {
  # FILL IN
}
```

b.  Make a time plot of the log absolute difference between `wn` and `wn_hat`.

c.  What consequence does this have for truncated forecasts?

d.  Compute the truncated forecast for $X_{53}$.

## 6. ACF, PACF, and BLPs

Let $(X_t)$ be a mean zero stationary process with the following autocovariance values:

$$
\gamma_X(0) = 2, \gamma_X(1) = 1.4, \gamma_X(2) = 0.6, \gamma_X(3) = 0.4, \gamma_X(4) = 0.2.
$$

a.  Can $(X_t)$ be an MA(2) process? Explain why or why not.

b.  Can $(X_t)$ be an AR(1) process? Explain why or why not.

c.  What is the best linear predictor $\hat X_4$ for $X_4$ given only $X_3 = 2$?

d.  Using the notation in part c), what is the variance of $X_4 - \hat X_4$?

e.  What is the best linear predictor $\hat X_4$ for $X_4$ given only $X_2 = 2$?

f.  Using the notation in part e), what is the variance of $X_4 - \hat X_4$?

g.  Let $\alpha_X$ denote the partial autocorrelation function of $(X_t)$. What is $\alpha_X(1)$?

h.  What is $\alpha_X(3)$?
